---
title: "Parkinson's Disease: EFA ja Classification Analysis"
subtitle: "Practical Work for DATA.STAT.450 Monimuuttujamenetelmät"
author: Aada Kerkkonen
output:
  pdf_document: default
  html_document: default
date: "22.2.2025"
fontsize: 12pt
bibliography: references.bib
biblio-style: "apalike"
link-citations: true
lang: en
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Table of Contents

1.  [Introduction](#introduction)
2.  [Data](#data)
3.  [Methodology](#methodology)
    1.  [Exploratory Factory Analysis](#efa)
    2.  [Classification Analysis](#classification)
4.  [Results](#results)
    1.  [EFA](#efa2)
    2.  [Classification Analysis](#classification2)
5.  [Conclusion](#conclusion)

\newpage

## Introduction <a name="introduction"></a> {#introduction}

Parkinson's disease (PD) is a neurodegenerative disease that causes
tremor, stiffness and other motor issues. As the disorder progresses,
the patient may also experience cognitive decline: mental health issues,
hallucinations and issues with attentiveness as well as speech and other
auditory issues. For example, people with Parkinson's disease tend to
speak monotonously, sometimes stuttering or forgetting words
[@MichaelJFox_Speech].

This project will analyse voice-related features in PD using Exploratory
Factory Analysis (EFA) and Classification Analysis. The former will be
used in hopes of discovering latent variables among the data and
reducing redundancy within the original dataset. These results will then
be used as input for the Classification Analysis. The aim is to classify
the observations into two groups: diagnosed with Parkinson's and not.

## Data <a name="data"></a> {#data}

The data for this project consists of 195 observations of different
biomedical voice measurements of 31 individuals [@parkinsons_174]. Out
of those 31, 23 are diagnosed with Parkinson's disease with time since
diagnosis ranging between 0 and 28 years. The data was originally used
for a study classifying people (46-85yo) with Parkinson's from those
without the disease by detecting dysphonia [@article]. Different
phonations were recorded from the subjects, ranging between a second and
36 seconds in length. To assure that the measures would be comparable to
those of other studies in the area, the researchers utilised a computer
program called MVDP (Multidimensional Voice Program Analysis).

The data consists of 24 variables: one for identification of the subject
(name), a boolean variable for the diagnosis (status) while the rest of
the variables are measurements of the phonations. These can be divided
into six categories:

1.  Frequency-Related measures in Hertz (relating to the pitch of the
    voice)
    -   MDVP:Fo - the average vocal fundamental frequency

    -   MDVP:Fhi - maximum vocal fundamental frequency

    -   MDVP:Flo - minimum vocal fundamental frequency
2.  Variation in Frequency (jitter, i.e. variance in timing and
    stability of vocal vibrations)
    -   MDVP:Jitter(%)

    -   MDVP:Jitter(abs)

    -   Jitter:DDP - difference between cycles

    -   MDVP:RAP - Relative Amplitude Perturbation

    -   MDVP:PPQ - Period Perturbation Quetient

        $\rightarrow$ High jitter value indicates unstable voice
        quality: can be a symptom of PD
3.  Amplitude Variation (shimmer, i.e. variations in the loudness of
    vocal vibrations)
    -   MDVP:Shimmer

    -   MDVP:Shimmer(dB)

    -   Shimmer:DDA - difference between cycles

    -   Shimmer:APQ3 - 3 point Amplitude Perturbation Quetient

    -   Shimmer:APQ5 - 5 point Amplitude Perturbation Quetient

    -   MDVP:APQ - 11 point Amplitude Perturbation Quetient

        $\rightarrow$ Higher shimmer value generally indicates
        irregulaties in loudness, i.e. can be a symptom of PD
4.  Noise-To-Harmonics (noise to tonal components, indicating the
    clarity of the sound)
    -   NHR - Noise To Harmonics Ratio

    -   HNR - Harmonics to Noise Ratio

        $\rightarrow$ A high NHR and low HNR can indicate PD
5.  Nonlinear Measures of Fundamental Frequency Variation
    -   spread1

    -   spread2

    -   PPE (Pitch Period Entropy, stability of speech)

        $\rightarrow$ Higher values can indicate instability in the
        speech, i.e. PD
6.  Other Variables (structure of the speech)
    -   RPDE - randomness of pitch patterns

    -   D2 - complexity of speech signal

    -   DFA - signal fractal scaling component

        $\rightarrow$ Can indicate chaotic structure in the speech, i.e.
        PD

Because the dataset consists of many different measurements of e.g.
jitter and shimmer, their respective correlations were analysed to infer
whether we could already reduce redundant variables as well as the
dimensionality of the data. Additionally, status is converted into a
factor.

```{r}
data <- read.table("parkinsons.data", sep = ",", header = TRUE)

# converting status into factor
data$status <- as.factor(data$status)

# jitter variables
cor(data[, c(5:9)])
```

As can be seen, all the variables are very highly correlated. Because of
this, only one of the jitter variables was kept for analysis:
MDVP.Jitter. This variable was chosen because the mean correlation in
its column is the highest. Therefore, all the jitter variables correlate
the most with MDVP.Jitter.

The same procedure was repeated for the shimmer variables:

```{r}
# shimmer variables
cor(data[, c(10:15)])
```

We can see that the variables are again very highly correlated. Taking
the variable with the highest mean: MDVP.Shimmer. Now we can remove the
redundant variables from the dataset and create another dataset that
only contains numerical values for EFA. Therefore, the name and status
columns are removed from that set.

```{r}
# removing redundant variables
filtered_data = data[, c(-(6:9), -(11:15))]
test_data = filtered_data[, c(-1, -9)]
dim(test_data)
```

We are left with 13 variables and 195 observations.

## Methodology <a name="methodology"></a> {#methodology}

The methods in this project consist of Exploratory Factory Analysis and
Classification Analysis. The information in this section is taken from
the book *Methods of Multivariate Analysis* [@alma9910683361405973] and
lecture slides.

### Exploratory Factory Analysis <a name="EFA"></a> {#efa}

EFA tries to find underlying constructs, factors, that can generate the
original variables in the data set through their correlations or
covariances. The aim is to reduce the redundancy among the dataset by
generating fewer factors than variables in the original set and express
$y_i$ as linear combinations of these factors. The model can be
expressed as

$$
\mathbf{y} - \mathbf{\mu} = \Lambda \mathbf{f}+\epsilon
$$

where

-   $y$ is the observed data

-   $\mu$ is the mean vector of the observed data

-   $\Lambda$ contains the loadings that show how each $y_i$ depends on
    the factors

-   $f$ is the vector of the underlying constructs, factors

-   $\epsilon$ is the random error

We can estimate the loadings through two main methods: Principal Factor
Analysis or Maximum Likelihood Factor Analysis. This project uses the
Maximum Likelihood method (MLE) because it can be used in hypothesis
testing for the covariance structure

$$
 H_0: \Sigma = \Lambda \Lambda'+ \Psi
$$

where

-   $\Lambda$ is as defined above

-   $\Psi$ is $cov(\epsilon)$ where the variances are of each
    $\epsilon_i$ are on the diagonal and every other cell is 0.

The p-value corresponding to this hypothesis testing can be found in the
printout. Because the p-value is the probability of a perfect fit
between the source data and number of factors, we want a larger number
[@visualstudiomagazine2017].

The result of EFA is not unique, and can be multiplied by an orthogonal
matrix $\bf{G}$ ($\bf{GG'=I}$) to rotate it and improve interpretation
or structure as follows

$$
\mathbf{y} = \Lambda \mathbf{G G' f} + \epsilon = \Lambda_* \mathbf{f}_*+ \epsilon
$$

The process of finding the best possible $\bf{G}$ is called rotation.
There are different methods of that will be tried in order to find the
best rotation for this dataset. In addition, we need to determine how
many factors to use in the model. Both of these will be assessed by the
cumulative variance; we want to use the model with the highest
cumulative variance.

In R, the analysis can be performed with the function *factanal()* that
takes in either the dataset, covariance or correlation matrix, the
number of factors, the method chosen for the estimation of loadings and
rotation. Additionally, we include the parameter *scores* to obtain the
factor scores used for the classification analysis and this result was
saved in a separate variable. These scores come from the conditional
distribution of $\bf{f}$ given $\bf{y}$.

In this project, the whole dataset was used as input for the model since
it is rather small. The *factanal* call was repeated for different
rotations and number of factors to obtain the highest cumulative
variance.

```{r}
results_efa <- factanal(test_data, factors = 6, method = "mle",
  rotation = "varimax", scores = "regression", n.obs = 195)

factor_scores <- results_efa$scores
```

### Classification Analysis <a name="classification"></a> {#classification}

The goal of classification analysis is to describe group separation
through discriminant functions (discriminant analysis) and then assign
observations to these groups (classification). In discriminant analysis,
a linear combination transforms each observation vector into a scalar
after which the means of the two groups are calculated.

$$
z = \mathbf{a} ' y \text{ and } \overline{z}_i= \mathbf{a}' \overline{y}_{i}
$$

We seek to find $\mathbf{a}$ that maximises standardizes the difference
between $\overline{z}_1$ and $\overline{z}_2$. The maximum is known to
occur at

$$
\mathbf{a}=S_{pl}^{-1}(\overline{y}_1 - \overline{y}_2)
$$

where

$$
S_{pl} = \frac{1}{n_1+n_2-2}((n_1-1)S_1+(n_2-1)S_2).
$$

Then to classify the observations into the two groups, we calculate the
discriminant scores of each observation and then the means of the groups
as defined above. To make the classification, we check whether the $z$
is closer to $\overline{z}_1$ or $\overline{z}_2$.

$$
z > \frac{1}{2}(\overline{z}_1+\overline{z}_2)
$$

If $z$ is larger as shown above, it belongs to group 1 and vice versa.

Additionally, because we want to test how the classification analysis
actually performs on unseen data, we split the data into test and
training groups. In this project, the model was trained with 80% of the
data and tested with the rest. In R, the package *mclust* provides a
function to perform both discriminant and classification analysis. The
parameters of *MclustDA* are the training set and the classes of those
observations, the type of the model and method of discriminant analysis.
Here we have chosen EDDA (Eigenvalue Decomposition Discriminant
Analysis) as model type and VVV (quadratic discriminant analysis) as the
special case. VVV was chosen because it assumes that each class have
their own covariance matrices, which is beneficial here as the different
groups may have very different covariances.

```{r}
library(mclust)

# splitting the data 
classification_data <- data.frame(factor_scores, status = filtered_data$status)
nc <- ncol(classification_data)
nr <- nrow(classification_data)

# seed set for reproducibility 
set.seed(42)
train_indices <- sample(1:nr, size = 0.8 * nr)

train <- classification_data[train_indices, -nc]  
train_class <- as.factor(classification_data$status[train_indices])

test <- classification_data[-train_indices, -nc]  
test_class <- as.factor(classification_data$status[-train_indices])

# producing model
results_clas <- MclustDA(train, train_class, 
                         modelType = "EDDA", modelNames = "VVV")
```

## Results <a name="results"></a> {#results}

### EFA <a name="EFA2"></a> {#efa2}

Let's first look through the results of the Exploratory Factory
Analysis. The chosen rotation was varimax and the number of factors in
the final model 6. This reached a cumulative variance between all the 6
factors of 80.5% and it was the highest obtained with different numbers
of factors and other rotations (promax and none)

```{r}
results_efa
```

Let's interpret the results. Uniqueness corresponds to $\Psi$,
proportion that is not explained by the variables, and $1 - \Psi$ gives
the communality: the proportion of the variance that the variable
contributed to the factors [@visualstudiomagazine2017]. Therefore, we
can see that the variables with high uniqueness, such as MDVP.Fhi.Hz,
MDVP.Flo.Hz and spread2, are poorly explained by the 6 factors. For
example, only 23.9% of the variation in MDVP.Fhi.Hz was retained with
the factors. In contrast, the factors, MDVP.Fo.Hz, MDVP.Jitter, NHR,
HNR, DFA, spread1, D2 and PPE, with low uniqueness are well explained by
the factors in the model. For example, 99.5% of MDVP.Fo.Hz is explained
with the factors. Therefore, the model is not capturing all the
variation in the variables and is thus not a perfect fit for the data.

As mentioned in the methodology section, loadings represent how each
variable depends on the factors. We can see that the first factor
accounts largely for MDVP.Jitter, MDVP.Shimmer, NHR and HNR (in the
opposite direction of NHR). This indicates that the first factor
captures a lot of the variability in the sound wave: amplitude and
period of the wave and the clarity of the sound. The second factor has
the highest scores for spread1, spread2 and PPE which means that the
factor accounts for the variation in frequency and thus the stability
and regularity of the speech. The third factor has the highest scores
for MDVP.Fo.Hz and MDVP.Flo.Hz, so the factor accounts frequency (pitch)
of voice. For factor 4 and 5 the highest values are for DFA (factor 4)
and D2 (factor 5). Therefore, these two factors largely account for the
complexity in speech. Finally, factor 6 has largest scores in HNR and
RPDE which could be interpreted as relating to the randomness of pitch
and the amount of noise. All together, we can see that the factors
identify similarities in the measurements that closely resemble those
groups explained in the section Data. The proportion of variance
explains how each factor contributes to the cumulative variance of the
model. We can see that even the final factor adds almost 8% to the
cumulative proportion of variance explained.

Finally, because we used MLE, the printout also includes a p-value.
Because this p-value is so small, we cannot reject the null hypothesis.
This means that the model does not perfectly fit the data and that more
factors could be needed.

### Classification Analysis <a name="classification2"></a> {#classification2}

Moving on to the classification analysis based on the factor scores
obtained in EFA.

```{r}
summary(results_clas, newdata = test, newclass = test_class)
```

The results indicate that the training set had 156 observations of which
119 were of observations of subjects diagnosed with PD. After performing
discriminant analysis based on the training set, the parameters found
were used for classifying the training set. The training confusion
matrix shows how the model classified the observations. Out of 37
observations that belonged to subjects without PD, the model predicted
29 (78.4%) correctly. In comparison, out of 119 with PD, the model
predicted 105 (88.2%) correctly. This gives the classification error of
14.1% where lower scores indicate better performance. Therefore, there
is a lot of room for improvement in the score. This score could indicate
that the data in itself has a lot of variety. This would be a reasonable
interpretation because auditory issues in PD appear as the disease
progresses and the time since diagnosis between subjects in the dataset
varied between 0 and 28 years.

With unseen data, the classification error is higher, 20.5%. This
indicates that the model is slightly overfitting the training data and
does not perform as well with new data. The model successfully
classified 9/11 (81.8%) correctly into the group of subjects without PD
and 22/28 (78.6%) into the PD group. The brier score supports our
findings as it is higher for the unseen data.

## Conclusion<a name="conclusion"></a> {#conclusion}

The project aimed to identify latent variables in Parkinson's data
through EFA and use these results as input in Classification Analysis to
classify observations into those with Parkinson's disease and those
without.

The EFA model successfully accounted for 80.5% of the variance in the
dataset with 6 factors. We could see that similar variables were grouped
together by the factors where the first factor highly accounted for
jitter, shimmer and noise in the voice, second and third to pitch and
frequency, fourth and fifth to complexity and finally the sixth factor
HNR so the amount of noise in the voice, indicating speech degradation.
While the proportion of variance explained by the model is rather high,
the low p-value and uniqueness scores indicate that the model was not
the perfect fit. I could not add more factors to improve the performance
of the model, so alternative methods are needed to address this. For
example, the data could be analysed more carefully with an expert in PD
and auditory problems in PD before EFA to further reduce variables and
identify the necessary ones. Although plenty of research was done, I was
not able to perfectly grasp the meanings and contributions of all the
variables in the dataset.

Classification Analysis successfully classified 78.6% of the subjects
with PD correctly and 81.8% for those without. While this is a
satisfactory score, it leaves room for a lot of improvement and suggests
overfitting the training data. Because the input for the Classification
Analysis was the factor score found through EFA, the classification
model did not attain all the information of the data. For example, some
features that would have improved the performance may have been lost.
Furthermore, since the time since diagnosis for those with PD varied
between 0 and 28 years, that could also account largely for the
variability and be a reason why both the factors and the classification
analysis where not able to perform perfectly. This could be included in
the dataset and thus provide better results for both EFA and
classification analysis. Additionally, ovefitting could be addressed by
using k-fold validation in place of just separating the data into
training and test sets.

In conclusion, the models performed reasonably well but will need
adjustments for better performance. These include contextualising the
variables and their meaning in the data set, adding a variable for time
since diagnosis and using k-fold validation to avoid overfitting.

# Sources {#sources}
